Decompositions of congruence subgroups of Chevalley groups

### 2015-10-15 paper.tex
Luzgarev comments + spacing in latin corrected

\section{Introduction}\label{sec:intro}
%Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots.
Let $\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over arbitarary commutative ring $R$ and let $\E(\Phi, R)$ be its \emph{elementary subgroup}
i.\, e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (aka \emph{principal congruence subgroup}) of level $I$, 
i.\, e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

@@

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z}(\Phi) = \{ z_\alpha(s, \xi) \mid s\in I, \ \xi \in R, \ \alpha \in \Phi\}.$

@@

The purpose of this paper is two-fold. 
First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
Secondly, we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.
%To be more specific, we study relative analogues of the following three kinds of decompositions:
%\begin{itemize}
%\item \emph{Gauss decomposition, } i.\, e. a decomposition of the form $\E(\Phi, R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup, 
%      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
%\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
%      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
%\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$, 
%      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
%\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

Our first result is a relative version of Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley $\E(\Phi, R, I)$ group admits the following decomposition
\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
\end{thm}
In the special case when $I$ is contained in the Jacobson radical of $R$ \cref{thm:srRI1} boils down to~\cite[Proposition~2.3]{Abe76}. 
We also note that the absolute case (i.\, e. $I=R$) of the theorem was treated by the second named author in~\cite{Sm12}.

@@

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\, e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
%The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $x_\alpha(\xi)$ has been extensively studied in a number of works (see e.\, g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
%Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\, g. localizations of Dedeking rings of arithmetic type.
%For example, W.~van der Kallen has shown that the width of $\SL(3, R)$ is infinite already for a Euclidean ring $R=\mathbb{C}[t]$.

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\E(\Phi, R)$ in terms of Tits generators and is, thus, a quantitative analogue of \cref{prop:Stepanov}
and a relative analogue of the results of~\cite{CK83, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}.
\begin{thm}\label{thm:width} Let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $R=\mathcal{O}_S$ is a Dedekind ring of arithmetic type posessing a real embedding and $\Phi$ is classical of rank $\geqslant2$, then 
$W(\G(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\item If $\sr(I) = 1$ and $\Phi$ is arbitrary irreducible root system then 
\[W(\E(\Phi, R, I), \mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1;\]
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime number $p$, then under the assumption of the Generalized Riemann Hypothesis one has
\begin{alignat*}{2}
& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell, \rC_\ell, \\
& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell, \rD_\ell;
\end{alignat*}
\end{enumerate}
\end{thm}
The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{TavgenThesis}), 
while for the second and the third assertion we give a direct proof.
Compare \cref{thm:width} with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.

@@

In~\cite{V13} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, in the context of \cref{thm:srRI1} one can immediately obtain the decomposition of $\E(\Phi, R, I)$ with $3|\Phi^+|$ factors of type $\rA_1$.
Our next result further extends this to even larger class of rings (which e.\, g. includes ideals of all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system satisfying the following assumptions.
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \rk(\Phi) - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\rk(\Phi).
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its (regularly embedded) subgroups isomorphic to $\SL_2(R, I)$.
\end{thm}

@@

\section{Preliminaries}\label{sec:prelim}
For any collection of subsets $H_1, \ldots, H_n$ of a group $G$ we denote by $H_1\ldots H_n$ their Minkowski set-product, 
i.\, e. the set consisting of arbitrary products $h_1\ldots h_n$ of elements $h_i\in H_i$. In particular, the equality
$G = H_1\cdot\ldots\cdot H_n$ means that every element $g\in G$ can be presented as a product $h_1\ldots h_n$ for $h_i\in H_i$.

\subsection{Root systems}\label{sec:rootsys}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots $\Pi=\{\alpha_1, \ldots, \alpha_l\}$.
We use conventional numbering of basis vectors of $\Pi$ which follows Bourbaki (cf.~\cite[Table~1]{PSV98}).
Denote by $\Phi^+$ and $\Phi^-$ the subsets of positive and negative roots with respect to $\Pi$.
Let $\alpha_\mathrm{max}$ denote the maximal root of $\Phi$.
For a root $\alpha\in\Phi$ we denote by $m_i(\alpha)$ the $i$-th component of $\alpha$ in $\Pi$, i.\, e. $\alpha = \sum_{i=1}^n m_i(\alpha) \alpha_i$.

@@

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
Let $\varpi_1, \ldots, \varpi_\ell$ be the \emph{fundamental weights }of $\Phi$, i.\, e. vectors
defined by $\langle\varpi_i, \alpha_j^\vee\rangle=\delta_{i, j}, $ where $\alpha^\vee = \frac{\alpha}{2(\alpha, \alpha)}$.

Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all reflections $\sigma_\alpha$ where $\sigma_\alpha(\beta)=\beta-\langle\alpha, \beta \rangle\cdot \alpha$.
For a reductive subset $\Delta\subseteq \Phi$ denote by $W(\Delta)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in\Delta$.

Let $J\subseteq \Pi$ be a subset of simple roots. 
For a root $\beta = \sum m_i(\beta)\alpha_i \in \Phi$ define \emph{$J$-shape} of $\beta$ (denoted $\shape(J, \beta)$) as $\sum_{i\in J} m_i(\beta) \alpha_i$.
Set
\begin{align*}
& \Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\}, \\
& \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha) \in \mathbb{Z}_{\geqslant0} \Phi^\pm \}, \\
& S_J^\pm = \Delta_J \sqcup \Sigma_J^\pm.
\end{align*}
Clearly, $\Delta_J$ is a reductive subset, while $S^\pm_J$ and $\Sigma^\pm_J$ are parabolic and special subsets, respectively.
For two disjoint subsets $I, J\subseteq \Pi$ one has 
\[ \Sigma^\pm_{I \cup J} = \Sigma^\pm_I\cup\Sigma^\pm_J, \quad \Delta_{I\cup J} = \Delta_I \cap \Delta_J. \]
We tend to omit curly braces in the above notations in the case when $J$ is a one or two element set, e.\, g. $\Delta_k=\Delta_{\{k\}}$ and $\Sigma_{i, j}^\pm=\Sigma_{\{i, j\}}^\pm$, etc.

@@

For an additive subset $I\subseteq R$ we denote by $X_\alpha(I)$ the corresponding \emph{root subgroup} of level $I$, i.\, e. the subgroup consisting of all elementary root unipotents $x_\alpha(\xi)$, $\xi\in I$.

@@

For $\varepsilon\in R^*$ set $w_\alpha(\varepsilon) = x_\alpha(\varepsilon) x_{-\alpha}(-\varepsilon^{-1}) x_{\alpha}(\varepsilon).$
If $\rk(\Phi)\geqslant 2$ the following relation is a consequence of \eqref{rel:add}--\eqref{rel:CCF}:
\begin{equation}\label{rel:R3}
w_\alpha(\varepsilon) x_{\beta}(\xi) w_\alpha(\varepsilon)^{-1} =
x_{\sigma\ssub{\alpha}\beta} \left(\eta_{\alpha, \beta}\cdot\varepsilon^{-\langle\beta, \alpha \rangle}\xi\right), \quad \varepsilon\in R^*, \ \xi\in R.
\end{equation}
where $\eta_{\alpha, \beta}=\pm 1$. The coefficients $\eta_{\alpha, \beta}$ can be expressed through structure constants (cf.~\cite[\S13]{VP}).
For a reductive subset $\Delta \subseteq \Phi$ denote by $\widetilde{W}(\Delta)$ the \emph{extended Weyl group}, i.\, e. the subgroup of $\E(\Phi, R)$ generated by all $w_{\alpha}(1)$, $\alpha \in \Delta$.
\begin{lemma} \label{lemma:weylfacts} Let $\Phi$ be an irreducible root system and let $I$ be an ideal of $R$. 
\begin{lemlist}
\item \label{item-trans1} For every roots $\alpha, \beta \in \Phi$ of the same length there exists $w \in \widetilde{W}(\Phi)$ such that $X_{\alpha}(I)^w = X_\beta(I)$.
\item \label{item-egen} The elementary group $\E(\Phi, R)$  is generated as an abstract group by root subgroups $X_{\alpha}(R)$, $\alpha \in \Pi \cup -\Pi$.
\item \label{item-trans2} Assume that $\alpha_s\in \Pi$ is a simple root of $\Phi$ such that $m_s(\alpha_\mathrm{max}) = 1$. 
                          Then for $\alpha, \beta \in \Sigma^\pm_s$ of the same length there exists $w\in \widetilde{W}(\Delta_s)$ such that $X_\alpha(I)^w = X_\beta(I)$. \end{lemlist}
\end{lemma}
\begin{proof}
The first statement of the lemma follows from identity~\eqref{rel:R3} and the well-known fact that the Weyl group $W(\Phi)$ acts transitively on the set of roots of the same length.

@@

A representation $\pi$ is called \emph{basic} if for every pair of non-zero weights $\lambda, \mu \in \Lambda^*(\pi)$ such that $\lambda - \mu \in \Pi$ one has $\sigma_\alpha(\lambda) = \mu$.
Obviously, this is equivalent to saying that $W(\Phi)$ acts transitively on $\Lambda(\pi)^*$.

The irreducible representations of Chevalley groups used in these paper form a sublist of~\cite[Table~2]{PSV98}.
More specifically, all of them are \emph{fundamental} (i.\, e. their highest weight is fundamental) and basic.
We widely employ the technique of \emph{weight diagrams} (see~\cite[\S~2]{PSV98}) to help visualize the structure of these representations.
In the simplest case of a \emph{microweight} representation $\pi$ (i.\, e. $\pi$ such that $V_0 = 0$)
the weight diagram of $(V, \pi)$ is just an undirected graph whose vertices are in one-to-one correspondence with the elements of $\Lambda(\pi)$.
Two vertices $\lambda$, $\mu$ of the diagram are joined by a bond marked $i$ iff $\lambda-\mu = \alpha_i$.

>>>>

\section{Introduction}\label{sec:intro}
%Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots.
Let $\G(\Phi, R)$ be the simply-connected Chevalley group of type $\Phi$ over arbitrary commutative ring $R$ and let $\E(\Phi, R)$ be its \emph{elementary subgroup}
i.\,e., the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$, see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (aka \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

@@

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z}(\Phi)$.

Here for a subset of roots $S\subseteq\Phi$ we denote by $\mathcal{Z}(S)$ the set
\[ \mathcal{Z}(S) = \{ z_\alpha(s, \xi) \mid s\in I, \ \xi \in R, \ \alpha \in \Phi\}.\]

@@

The purpose of this paper is two-fold. 
First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
Secondly, we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.
%To be more specific, we study relative analogues of the following three kinds of decompositions:
%\begin{itemize}
%\item \emph{Gauss decomposition, } i.\,e. a decomposition of the form $\E(\Phi, R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup, 
%      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
%\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
%      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
%\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$, 
%      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
%\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

Our first result is a relative version of Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley group $\E(\Phi, R, I)$ admits the following decomposition
\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
\end{thm}
In the special case when $I$ is contained in the Jacobson radical of $R$, \cref{thm:srRI1} boils down to~\cite[Proposition~2.3]{Abe76}. 
We also note that the absolute case (i.\,e. $I=R$) of the theorem was treated by the second named author in~\cite{Sm12}.

@@

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
%The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $x_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
%Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
%For example, W.~van der Kallen has shown that the width of $\SL(3, R)$ is infinite already for a Euclidean ring $R=\mathbb{C}[t]$.

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\E(\Phi, R)$ in terms of Tits generators and is, thus, a quantitative analogue of \cref{prop:Stepanov}
and a relative analogue of the results of~\cite{CK83, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}.
\begin{thm}\label{thm:width} Let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $R=\mathcal{O}_S$ is a Dedekind ring of arithmetic type posessing a real embedding and $\Phi$ is classical of rank $\geqslant2$, then 
$W(\G(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\item If $\sr(I) = 1$ and $\Phi$ is arbitrary irreducible root system, then 
\[W(\E(\Phi, R, I), \mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1;\]
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime number $p$, then under the assumption of the Generalized Riemann Hypothesis one has
\begin{alignat*}{2}
& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell, \rC_\ell, \\
& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell, \rD_\ell;
\end{alignat*}
\end{enumerate}
\end{thm}
The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{TavgenThesis}), 
while for the second and the third assertion we give a direct proof.
Compare \cref{thm:width} with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.

@@

In~\cite{V13} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, in the context of \cref{thm:srRI1} one can immediately obtain the decomposition of $\E(\Phi, R, I)$ with $3|\Phi^+|$ factors of type $\rA_1$.
Our next result further extends this to even larger class of rings (which e.\,g. includes ideals of all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system satisfying the following assumptions.
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \rk(\Phi) - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\rk(\Phi).
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its (regularly embedded) subgroups isomorphic to $\SL_2(R, I)$.
\end{thm}

@@

\section{Preliminaries}\label{sec:prelim}
For any collection of subsets $H_1, \ldots, H_n$ of a group $G$ we denote by $H_1\ldots H_n$ their Minkowski set-product, 
i.\,e. the set consisting of arbitrary products $h_1\ldots h_n$ of elements $h_i\in H_i$. In particular, the equality
$G = H_1\cdot\ldots\cdot H_n$ means that every element $g\in G$ can be presented as a product $h_1\ldots h_n$ for $h_i\in H_i$.

\subsection{Root systems}\label{sec:rootsys}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots $\Pi=\{\alpha_1, \ldots, \alpha_l\}$.
We use conventional numbering of basis vectors of $\Pi$ which follows Bourbaki (cf.~\cite[Table~1]{PSV98}).
Denote by $\Phi^+$ and $\Phi^-$ the subsets of positive and negative roots with respect to $\Pi$.
Let $\alpha_\mathrm{max}$ denote the maximal root of $\Phi$.
For a root $\alpha\in\Phi$ we denote by $m_i(\alpha)$ the $i$-th component of $\alpha$ in $\Pi$, i.\,e. $\alpha = \sum_{i=1}^n m_i(\alpha) \alpha_i$.

@@

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
Let $\varpi_1, \ldots, \varpi_\ell$ be the \emph{fundamental weights }of $\Phi$, i.\,e. vectors
defined by $\langle\varpi_i, \alpha_j^\vee\rangle=\delta_{i, j}, $ where $\alpha^\vee = \frac{\alpha}{2(\alpha, \alpha)}$.

Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all reflections $\sigma_\alpha$ where $\sigma_\alpha(\beta)=\beta-\langle\beta, \alpha \rangle\cdot \alpha$.
For a reductive subset $\Delta\subseteq \Phi$ denote by $W(\Delta)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in\Delta$.

Let $J\subseteq \Pi$ be a subset of simple roots. 
For a root $\beta = \sum m_i(\beta)\alpha_i \in \Phi$ define \emph{$J$-shape} of $\beta$ (denoted $\shape(J, \beta)$) as $\sum_{i\in J} m_i(\beta) \alpha_i$.
Set
\begin{align*}
& \Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\}, \\
& \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha) \in \mathbb{Z}_{\geqslant0} \Phi^\pm \}, \\
& S_J^\pm = \Delta_J \sqcup \Sigma_J^\pm.
\end{align*}
Clearly, $\Delta_J$ is a reductive subset, while $S^\pm_J$ and $\Sigma^\pm_J$ are parabolic and special subsets, respectively.
For two disjoint subsets $I, J\subseteq \Pi$ one has 
\[ \Sigma^\pm_{I \cup J} = \Sigma^\pm_I\cup\Sigma^\pm_J, \quad \Delta_{I\cup J} = \Delta_I \cap \Delta_J. \]
We tend to omit curly braces in the above notations in the case when $J$ is a one or two element set, e.\,g. $\Delta_k=\Delta_{\{k\}}$ and $\Sigma_{i, j}^\pm=\Sigma_{\{i, j\}}^\pm$, etc.

@@

For an additive subset $I\subseteq R$ we denote by $X_\alpha(I)$ the corresponding \emph{root subgroup} of level $I$, i.\,e. the subgroup consisting of all elementary root unipotents $x_\alpha(\xi)$, $\xi\in I$.

@@

For $\varepsilon\in R^*$ set $w_\alpha(\varepsilon) = x_\alpha(\varepsilon) x_{-\alpha}(-\varepsilon^{-1}) x_{\alpha}(\varepsilon).$
If $\rk(\Phi)\geqslant 2$ the following relation is a consequence of \eqref{rel:add}--\eqref{rel:CCF}:
\begin{equation}\label{rel:R3}
w_\alpha(\varepsilon) x_{\beta}(\xi) w_\alpha(\varepsilon)^{-1} =
x_{\sigma\ssub{\alpha}\beta} \left(\eta_{\alpha, \beta}\cdot\varepsilon^{-\langle\beta, \alpha \rangle}\xi\right), \quad \varepsilon\in R^*, \ \xi\in R.
\end{equation}
where $\eta_{\alpha, \beta}=\pm 1$. The coefficients $\eta_{\alpha, \beta}$ can be expressed through structure constants (cf.~\cite[\S13]{VP}).
For a reductive subset $\Delta \subseteq \Phi$ denote by $\widetilde{W}(\Delta)$ the \emph{extended Weyl group}, i.\,e. the subgroup of $\E(\Phi, R)$ generated by all $w_{\alpha}(1)$, $\alpha \in \Delta$.
\begin{lemma} \label{lemma:weylfacts} Let $\Phi$ be an irreducible root system and let $I$ be an ideal of $R$. 
\begin{lemlist}
\item \label{item-trans1} For every roots $\alpha, \beta \in \Phi$ of the same length there exists $w \in \widetilde{W}(\Phi)$ such that $X_{\alpha}(I)^w = X_\beta(I)$.
\item \label{item-egen} The elementary group $\E(\Phi, R)$  is generated as an abstract group by root subgroups $X_{\alpha}(R)$, $\alpha \in \Pi \cup -\Pi$.
\item \label{item-trans2} Assume that $\alpha_s\in \Pi$ is a simple root of $\Phi$ such that $m_s(\alpha_\mathrm{max}) = 1$. 
                          Then for $\alpha, \beta \in \Sigma^\pm_s$ of the same length there exists $w\in \widetilde{W}(\Delta_s)$ such that $X_\alpha(I)^w = X_\beta(I)$. \end{lemlist}
\end{lemma}
\begin{proof}
The first statement of the lemma follows from identity~\eqref{rel:R3} and the well-known fact that the Weyl group $W(\Phi)$ acts transitively on the set of roots of the same length.

@@

A representation $\pi$ is called \emph{basic} if for every pair of non-zero weights $\lambda, \mu \in \Lambda^*(\pi)$ such that $\lambda - \mu \in \Pi$ one has $\sigma_\alpha(\lambda) = \mu$ for some $\alpha\in\Phi$.
Obviously, this is equivalent to saying that $W(\Phi)$ acts transitively on $\Lambda(\pi)^*$.

The irreducible representations of Chevalley groups used in these paper form a sublist of~\cite[Table~2]{PSV98}.
More specifically, all of them are \emph{fundamental} (i.\,e. their highest weight is fundamental) and basic.
We widely employ the technique of \emph{weight diagrams} (see~\cite[\S~2]{PSV98}) to help visualize the structure of these representations.
In the simplest case of a \emph{microweight} representation $\pi$ (i.\,e. $\pi$ is basic and $V_0 = 0$)
the weight diagram of $(V, \pi)$ is just an undirected graph whose vertices are in one-to-one correspondence with the elements of $\Lambda(\pi)$.
Two vertices $\lambda$, $\mu$ of the diagram are joined by a bond marked $i$ iff $\lambda-\mu = \alpha_i$.

### 2015-10-15 stability-conditions.tex
Luzgarev comments + spacing in latin corrected

An $I$-unimodular row $a=(a_1,\ldots a_{n+1})$ is called \emph{stable} if one can choose $b_1,\ldots,b_n\in I$ such that the row $(a_1+a_{n+1}b_1,\ldots,a_n+a_{n+1}b_n)$ is also $I$-unimodular. 
We say that $I$ satisfies stable range condition $\SR_n(I)$ if any $I$-unimodular row of length $n+1$ is stable.
It is easy to see that $\SR_n(I)$ implies $\SR_m(I)$ for any $m\geqslant n$. 
It is also clear that condition $\SR_n(I)$ does not depend on the choice of the ambient ring $R$.
By definition, the {\it stable rank} of $I$ (denoted $\sr(I)$) is the smallest natural $n$ such that $\SR_n(I)$ holds (we set $\sr(I)=\infty$ if no such $n$ exists).

>>>>

An $I$-unimodular row $a=(a_1,\ldots a_{n+1})$ is called \emph{stable} if one can choose $b_1,\ldots,b_n\in I$ such that the row $(a_1+a_{n+1}b_1,\ldots,a_n+a_{n+1}b_n)$ is also $I$-unimodular. 
We say that $I$ satisfies stable range condition $\SR_n(I)$ if any $I$-unimodular row of length $n+1$ is stable.
Vaserstein noted~\cite[Theorem~1]{Va71} that $\SR_n(I)$ implies $\SR_m(I)$ for any $m\geqslant n$. 
It is also clear that condition $\SR_n(I)$ does not depend on the choice of the ambient ring $R$.
By definition, the {\it stable rank} of $I$ (denoted $\sr(I)$) is the smallest natural number $n$ such that $\SR_n(I)$ holds (we set $\sr(I)=\infty$ if no such $n$ exists).
